\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}  % Adds to TOC

This project addresses the growing need for intelligent and automated systems to manage traffic violations, particularly speed limit breaches. It aims to leverage low-cost embedded technologies to detect speed violations, recognize license plate numbers, manage offender data, and automate the fine notification process. The primary objective is to automate speed violation detection and contribute to behavior change among drivers by consistently recording violations and applying penalties.The system uses an ultrasonic sensor for vehicle speed measurement, with an Arduino Uno handling signal processing and speed calibration. A Raspberry Pi, equipped with a camera module, captures images of speeding vehicles. License Plate Recognition (LPR) is achieved using OpenCV and EasyOCR, while SQLite is used for local data storage. Offender notifications are handled through an automated email system using SMTP. Features of the system include speed detection, reliable license plate recognition, and a basic fine system based on the severity of the violation. While the current implementation serves as an academic proof-of-concept, its real-world deployment could significantly enhance traffic law enforcement. It has the potential to reduce accidents caused by overspeeding, minimize corruption by eliminating the need for human traffic officers, and encourage responsible driving by maintaining a record of each driver’s violations.